### 2017-02-28 doc/manuscript.Rmd
Minor edits to manuscript as per Tony's suggestions

Word count: ~4200

@@

The interaction between TAG and insulin sensitivity is complex and has 
components of a feedback system [@Verges2015a]. Greater resistance to insulin in
both the liver and muscle may result in greater production of TAG and secretion
of lipoproteins that transport TAG [@Yu2012a]. Likewise, greater TAGFA may
contribute to metabolic dysfunction and lipotoxicity in various tissues, such as
the beta-cells, and thus continuing the cycle [@Verges2015a]. Given the
complexity and temporal nature of the relationship, long term studies with
multiple data collections are paramount to better understanding the underlying
biology and subsequent risk.

@@

Insulin sensitivity and beta-cell function indices were computed using the OGTT 
glucose and insulin data. Insulin sensitivity was assessed using the Insulin
Sensitivity Index (ISI) [@Matsuda1999] and HOMA2-%S [@Levy1998a] using the HOMA2
Calculator. HOMA largely reflects hepatic insulin resistance, while ISI
reflects whole-body insulin sensitivity [@AbdulGhani2007]. Beta-cell function
was assessed using the Insulinogenic Index [@Wareham1995] over HOMA-IR [@Matthews1985] (IGI/IR)
and the Insulin Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR
is a measure of the first phase of insulin secretion while ISSI-2 is analogous to
the disposition index (but is calculated using OGTT values). Each index has been validated
against gold standard measures [@Matthews1985; @Hermans1999a; @Matsuda1999; @Retnakaran2009].
Specific formulas for each OGTT-derived measure can be found in the 
Supplementary Material.

@@

While GEE accounts for the longitudinal design of the data, this approach is 
limited in that it cannot analyze the inherent multivariate nature of the 
composition of the TAGFA fraction. Therfore, to confirm the GEE results in a multivariate 
environment (i.e. all TAGFA analyzed collectively), partial least squares regression (PLS) was
used to identify the patterns of TAGFA composition against insulin sensitivity
and beta-cell function as outcome variables. For a detailed explanation of PLS
see the Supplemental Methods. Briefly, PLS is a technique that extracts latent
structures (clusters) underlying a set of predictor variables conditional on a
response variable(s) (i.e. the outcome variables). How accurately the clusters
within the TAGFA composition predict metabolic function is determined by using
cross-validation on the PLS models.

@@

`r cite_t('basic')` shows basic characteristics of the PROMISE cohort. The mean 
follow-up time was `r calc_followup_time()` years, where `r calc_pct_full_visits()`%
of participants attended all three visits. There were
`r female$npct` females and `r euro$npct` who had European-ancestry, with a mean
age in years of `r aide::ave_sd(ds0yr$Age)` and a mean BMI
of `r aide::ave_sd(ds0yr$BMI)` kg/m^2^. As expected from the study's eligibility criteria, the majority of
participants (`r fhd$npct`) had a family history of diabetes.
Between the baseline visit and the 6-year visit in this sample, insulin
sensitivity and beta-cell function measures had a significant median decline of
between `r outcome_chg$chg` (p`r outcome_chg$p` from GEE; n=`r outcome_chg$n`). There 
were `r dys$pre` and `r dys$dm` incident cases of pre-diabetes (IFG and IGT) and
diabetes, respectively, over the 6-years; these observations were excluded from GEE and PLS
analyses.

@@

Results from the unadjusted GEE model are shown in `r cite_f('gee_unadj')`) and 
for the adjusted GEE model in `r cite_f('gee_adj')`. The majority of
associations with beta-cell function measures were attenuated after full model
adjustment, while nearly all associations with insulin sensitivity remained
significant for both mol% and nmol/mL results. Subsequent analysis revealed that
the attenuation with beta-cell function was due primarily to adjustment for
waist.

In analyses using concentration values, nearly all TAGFA had a strong
negative association on HOMA2-%S and ISI (estimates ranging from
`r est1$outcome_b$rng[3]` and `r est1$outcome_b$rng[7]`, respectively), and a
few had strong negative associations with IGI/IR and ISSI-2 (estimates ranging 
from `r est1$outcome_b$rng[11]` and `r est1$outcome_b$rng[15]`, respectively).
In analyses using TAGFA mol% values, four TAGFA (14:0, 16:0, 14:1n-7, and 16:1n-7) had negative
associations with HOMA2-%S and ISI (between `r est1$outcome_b$rng[1]` and 
`r est1$outcome_b$rng[5]`%, respectively, lower insulin sensitivity for every SD
increase in the TAGFA), while several more TAGFA had positive associations with
HOMA2-%S and ISI (20:0, 18:1n-9, 20:1n-9, 22:1n-9, 18:2n-6, 20:2n-6, 20:4n-6,
and 22:5n-3) predicting between `r est1$outcome_b$rng[2]` and 
`r est1$outcome_b$rng[6]`%, respectively, higher insulin sensitivity for every
SD increase in the TAGFA. One TAGFA, 20:2n-6, had a very strong positive
association with the insulin sensitivity measures, with a `r est1$b_ci$est[1]`
to `r est1$b_ci$est[37]`% higher insulin sensitivity for every SD increase. 
Both clinically-measured TAG and total TAGFA concentration had very strong
negative associations with all outcome variables.

@@

To our knowledge, no longitudinal study to date has examined the role of the
composition of the TAGFA fraction on detailed OGTT-derived metabolic measures 
Two large prospective studies have been published that similarly examined TAGFA 
composition and diabetes outcomes. Rhee *et al* presented a nested case-control
analysis (n=189 cases and n=189 controls) within the Framingham offspring cohort
[@Rhee2011a], which found that subjects with a TAGFA composition characterized
by a lower carbon chain and fewer double bonds (e.g. 14:0, 16:0) had a higher
risk for diabetes after 12-years while those with a profile characterized by
higher carbon chain and more double bond TAGFA had a lower risk for diabetes.
A similar pattern of TAGFA was also associationed with HOMA-IR cross-sectionally
at the baseline visit. In addition, Lankinen *et al* reported on
a prospective cohort of males in Finland [@Lankinen2015a], for which 
TAGFA data were available for 831 participants after 6-years of follow-up. 
They found that most saturated fatty acids had negative , docosapentaenoic acid (22:5n-3),
eicosapentaenoic acid (20:5n-3), and arachidonic acid (20:4n-6), though they did
not find any associations with 7-tetradecenoic acid (14:1n-7) and palmitoleic
acid (16:1n-7). {{ fix up }}
There were some
similarities in results between the present analysis and the Finnish study,
Our study extends these findings by using multiple measurements
of metabolic function and as well as multivariate statistical approaches that
allowed us to identify a cluster of TAGFA.
In another study of a much
smaller (n=16) mostly female group [@Kotronen2009a], the authors report a
positive correlation between total esterified (of which TAG make up the
majority) 16:0, 16:1n-7, and 18:1n-9 with HOMA-IR, similar as our results except
for the 18:1n-9 finding.

@@

{{fix this}}
However, while this DNL cluster had a strong negative association
with insulin sensitivity, the magnitude of the association was slightly weaker
compared to clinically measured TAG.

>>>>

Word count: ~4000

@@

The interaction between TAG and insulin sensitivity is complex and has 
components of a feedback system [@Verges2015a]. Greater resistance to insulin in
both the liver and muscle may result in greater production of TAG and secretion
of lipoproteins that transport TAG [@Yu2012a]. Likewise, greater TAG may
contribute to metabolic dysfunction and lipotoxicity in various tissues, such as
the beta-cells, and thus continuing the cycle [@Verges2015a]. Given the
complexity and temporal nature of the relationship, long term studies with
multiple data collections are paramount to better understanding the underlying
biology and subsequent risk.

@@

Insulin sensitivity and beta-cell function indices were computed using the OGTT 
glucose and insulin data. Insulin sensitivity was assessed using the Insulin
Sensitivity Index (ISI) [@Matsuda1999] and HOMA2-%S [@Levy1998a] using the HOMA2
Calculator. HOMA largely reflects hepatic insulin sensitivity, while ISI
reflects whole-body insulin sensitivity [@AbdulGhani2007]. Beta-cell function
was assessed using the Insulinogenic Index [@Wareham1995] over HOMA-IR [@Matthews1985] (IGI/IR)
and the Insulin Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR
is a measure of the first phase of insulin secretion while ISSI-2 is analogous to
the disposition index (but is calculated using OGTT values). Each index has been validated
against gold standard measures [@Matthews1985; @Hermans1999a; @Matsuda1999; @Retnakaran2009].
Specific formulas for each OGTT-derived measure can be found in the 
Supplementary Material.

@@

While GEE accounts for the longitudinal design of the data, this approach is 
limited in that it cannot analyze the inherent multivariate nature of the 
composition of the TAGFA fraction. Therefore, to confirm the GEE results in a multivariate 
environment (i.e. all TAGFA analyzed collectively), partial least squares regression (PLS) was
used to identify the patterns of TAGFA composition against insulin sensitivity
and beta-cell function as outcome variables. For a detailed explanation of PLS
see the Supplemental Methods. Briefly, PLS is a technique that extracts latent
structures (clusters) underlying a set of predictor variables conditional on a
response variable(s) (i.e. the outcome variables). How accurately the clusters
within the TAGFA composition predict metabolic function is determined by using
cross-validation on the PLS models.

@@

`r cite_t('basic')` shows basic characteristics of the PROMISE cohort. The mean 
follow-up time was `r calc_followup_time()` years, where `r calc_pct_full_visits()`%
of participants attended all three visits. There were
`r female$npct` females and `r euro$npct` who had European-ancestry, with a mean
age in years of `r aide::ave_sd(ds0yr$Age)` and a mean BMI
of `r aide::ave_sd(ds0yr$BMI)` kg/m^2^. As expected from the study's eligibility
criteria, the majority of participants, n=`r fhd$npct`, had a family history of
diabetes. Between the baseline visit and the 6-year visit in this sample,
insulin sensitivity and beta-cell function measures had a significant median decline of
between `r outcome_chg$chg` (p`r outcome_chg$p` from GEE; n=`r outcome_chg$n`). There 
were `r dys$pre` and `r dys$dm` incident cases of pre-diabetes (IFG and IGT) and
diabetes, respectively, over the 6-years; these observations were excluded from GEE and PLS
analyses.

@@

Results from the unadjusted GEE model are shown in `r cite_f('gee_unadj')` and 
for the adjusted GEE model in `r cite_f('gee_adj')`. The majority of
associations with beta-cell function measures were attenuated after full model
adjustment, while nearly all associations with insulin sensitivity remained
significant for both mol% and nmol/mL results. Subsequent analysis revealed that
the attenuation with beta-cell function was due primarily to adjustment for
waist.

In analyses using concentration values, nearly all TAGFA had a strong
negative association on HOMA2-%S and ISI (estimates ranging from
`r est1$outcome_b$rng[3]` and `r est1$outcome_b$rng[7]`, respectively), and a
few had strong negative associations with IGI/IR and ISSI-2 (estimates ranging 
from `r est1$outcome_b$rng[11]` and `r est1$outcome_b$rng[15]`, respectively).
In analyses using TAGFA mol% values, four TAGFA (14:0, 16:0, 14:1n-7, and 16:1n-7) had negative
associations with HOMA2-%S and ISI (between `r est1$outcome_b$rng[1]` and 
`r est1$outcome_b$rng[5]`, respectively, lower insulin sensitivity for every SD
increase in the TAGFA), while several more TAGFA had positive associations with
HOMA2-%S and ISI (20:0, 18:1n-9, 20:1n-9, 22:1n-9, 18:2n-6, 20:2n-6, 20:4n-6,
and 22:5n-3) predicting between `r est1$outcome_b$rng[2]` and 
`r est1$outcome_b$rng[6]`%, respectively, higher insulin sensitivity for every
SD increase in the TAGFA. One TAGFA, 20:2n-6, had a very strong positive
association with the insulin sensitivity measures, with a `r est1$b_ci$est[1]`
to `r est1$b_ci$est[37]`% higher insulin sensitivity for every SD increase. 
Both clinically-measured TAG and total TAGFA concentration had very strong
negative associations with all outcome variables.

@@

To our knowledge, no longitudinal study to date has examined the role of the
composition of the TAGFA fraction on detailed OGTT-derived metabolic measures 
Two large prospective studies have been published that similarly examined TAGFA 
composition and diabetes outcomes. Rhee *et al* presented a nested case-control
analysis (n=189 cases and n=189 controls) within the Framingham offspring cohort
[@Rhee2011a], which found that subjects with a TAGFA composition characterized
by a lower carbon chain and fewer double bonds (e.g. 14:0, 16:0) had a higher
risk for diabetes after 12-years while those with a profile characterized by
higher carbon chain and more double bond TAGFA had a lower risk for diabetes.
A similar pattern of TAGFA was also associated with HOMA-IR cross-sectionally
at the baseline visit. In addition, Lankinen *et al* reported on
a prospective cohort of males in Finland [@Lankinen2015a], for which 
TAGFA data were available for 831 participants after 6-years of follow-up. In
their cohort, OGTT data was only available at the 6-year visit.
They cross-sectionally found that most saturated fatty acids had negative 
associations with insulin sensitivity and beta-cell function while
linoleic acid (18:2n-6), docosapentaenoic acid (22:5n-3), eicosapentaenoic acid
(20:5n-3), and arachidonic acid (20:4n-6) had positive associations with insulin
sensitivity. The magnitude of the associations were larger in the insulin
sensitivity results compared to the beta-cell function results, similar to what
we observed. Our study extends these findings by using multiple measurements
of metabolic function and as well as multivariate statistical approaches that
allowed us to identify clusters of TAGFA. In another study of a much
smaller (n=16) mostly female group [@Kotronen2009a], the authors report a
positive correlation between total esterified (of which TAG make up the
majority) 16:0, 16:1n-7, and 18:1n-9 with HOMA-IR, similar as our results except
for the 18:1n-9 finding.

@@
